---
title: 'Reproducible Research: Peer Assignment 1'
author: "Divine Miho"
date: "Wednesday, February 11, 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

Set the working directory

```{r}
setwd("~/Coursera/Reproducible_Research")
```

Load the required packages

```{r}
library(stats)
library(plyr)
library(lattice)
```

Load and manipulate the data

```{r}
dat <- read.csv("activity.csv")
```

Process/transform the data to a suitable format

```{r}
dat$date <- as.Date(dat$date, "%Y-%m-%d")
```

Summarize the dependent variable of interest, steps

```{r}
summary(dat$steps)
```

Total number of steps per day

```{r}
sum(dat$steps, na.rm=TRUE)
```

Histogram of the total number of steps taken each day

```{r}
totStps <-aggregate(dat$steps, by=list(dat$date), FUN=sum)
colnames(totStps) <- c("Date","total_steps")

h<-hist(totStps$total_steps, breaks=30, col="red", 
        xlab="Total Steps", 
        main="Histogram of Total Steps Taken Each Day")
```

Mean of the total steps taken each day

```{r}
mean(totStps$total_steps,na.rm = TRUE)
```

Median of the total steps taken each day

```{r}
median(totStps$total_steps,na.rm = TRUE)
```

Time series plot of five minute intervals and average number of steps taken

Get the data

```{r}
aveStps <-aggregate(dat$steps, by=list(dat$interval), FUN=mean, na.rm= TRUE)
colnames(aveStps) <- c("interval","ave_steps")
with(aveStps, plot(interval,ave_steps, type = "l",
                   main = "Average steps by time intervals",
                   xlab = "Time intervals (minutes)",
                   ylab = "Average steps across all days"))
```

The five minute interval with the maximum numer of steps, on average

```{r}
which.max(aveStps[,2])
maxInt <- aveStps[aveStps$ave_steps == max(aveStps$ave_steps),]
```

Interval with max average total steps is `r maxInt[[1]]`

Since the data has missing values, I will be doing some imputation

I am replacing the missing values with the average total steps by interval

Count the number of rows with missing values

```{r}
sum(is.na(dat$steps))
```

Imputation

```{r}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
dat2 <- ddply(dat, ~ interval, transform, steps = impute.mean(steps))
dat2 <- with(dat2, dat2[order(date,interval),])
```

Inorder to create a histogram, I need to aggregate steps by day

```{r}
totStps2 <-aggregate(dat2$steps, by=list(dat2$date), FUN=sum)
colnames(totStps2) <- c("Date","total_steps")
```

Histogram of the total number of steps taken each day

```{r}
h2<-hist(totStps2$total_steps, breaks=30, col="red", 
         xlab="Total Steps", 
         main="Histogram of Total Steps Taken Each Day")
```

I am plotting the two histograms together to see if the imputation
made any difference to the distribution

```{r}
hist(totStps$total_steps, col=rgb(1,0,0,0.5),breaks = 30,
     xlab="Total Steps", main="Histogram of Total Steps Taken Each Day")
hist(totStps2$total_steps, breaks = 30, col=rgb(0,0,1,0.5), add=T)
box()
```

Compare the means and medians with and without the imputation

```{r}
s1 <- summary(totStps2$total_steps)
m1 <- s1[[4]]
md1 <- s1[[3]]
s2 <- summary(totStps$total_steps)
m2 <- s2[[4]]
md2 <- s2[[3]]
```

cat("Before the imputation, the mean was", `r m1`,
    "and after imputation, the mean is", `r m2`)

cat("Before the imputation, the median was", `r md1`,
    "and after imputation, the median is", `r md2`)

Imputataion affected the median (lowered it) but had no impact on the mean

Now I examine difference in activity between weekedays and weekends

First I create new factor variable for weekend and weekday dates

```{r}
dat2$wknd <- as.factor(ifelse(weekdays(dat2$date) %in% c("Saturday","Sunday"),
                              "Weekend","Weekday")) 
```

Here is a panel plot comparing the total steps taken on the weekend vs. weekdays

```{r}
dat3 <- ddply(dat2, c("wknd", "interval"), summarise,
              total_steps = sum(steps))

xyplot(total_steps~interval|wknd,dat3,type='l', cex = 0.3, layout = c(1, 2),
       xlab = 'Interval', ylab = 'Total Steps',
       main = "Total steps by interval by weekday type")
```